---
title: "quick_analysis_forbradSep2024"
author: "Eddy Dowle"
date: "2024-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Overview

A quick and dirty analysis of some of brads data using vegan


```{r import, echo=TRUE}
#having a look at the files
library('tidyverse')
library('readxl')
library('vegan')

setwd("C:/Users/hrlexd/Dropbox/PlantAndFood (1)/B4BI/Analysis_Sep2024")
#im treating blanks as 0 I have discussed the difference between a null or NA and a 0 with brad and he thinks all blank cells should be 0
#insect visitation data
insects_2024<-read_excel('Flower visit data Eddy 2024_noblanks.xlsx', sheet = '2024 data',na='NA')
#I have fixed various issues with this file but there is still the double rows for Marr in it
head(insects_2024)
#meta data
#tidied up due to some doubles so use tidy version for metadata
meta_data_farm<-read.csv('Farm_meta_data_sep2024.csv')
head(meta_data_farm)

unique(insects_2024$Native_Plant_Species)
#brad wants to focus on
#Veronica_salicifolia hebe
#Leptospermum_scoparium kanuka
#Cordyline_australis cabbage tree
```

## Just looking at Veronica_salicifolia for now

```{r dataframe, echo=TRUE}
test<-insects_2024 %>% filter(Native_Plant_Species=="Veronica_salicifolia") %>% select(Insect_key_name,Timeperiod,Property_name,Tree_number,Insect_count)

unique(test$Insect_key_name)
unique(test$Timeperiod)
unique(test$Property_name)
unique(test$Insect_count)
unique(test$Tree_number)

```

## What do we want?
Diversity across farms, diversity morning afternoon midday?
Dversity across tree species within a farm? Could group plots by farm?

I think for the diversity plots it would be good to include the diversity across the three species side by side but will get feedback before going further

```{r cleanup, echo=TRUE}
#getting rid of the Marr dups by summing them (not sure if this is the way we should deal with them)
data_long <- test %>% 
  group_by(Property_name, Timeperiod,Tree_number,Insect_key_name) %>%
  summarise(Insect_count = sum(Insect_count))
#here is an example of the Marr dups:
#its not a straight dup as the counts vary across the dups so I dont know whats going on but Im just going to keep the sumed values, maybe two people did the survey and it got entered twice???? who knows
test %>% filter(Property_name=='Marr',Insect_key_name=='Honey bee', Timeperiod=='Morning')

#so we want a spreadsheet of:
#farm #time #rep #counts_spA #counts_spB etc
data_long_wide<-data_long %>% pivot_wider(names_from = Insect_key_name, values_from = Insect_count)

```

## Abundance plots
Need to recolour the species groupings, brad has given me the taxonomy file I just havent incorporated it yet

```{r abundance, echo=TRUE}
#raw abundance plots
data_long$Timeperiod <- factor(data_long$Timeperiod, levels=c("Morning", "Midday", "Afternoon"))

#legends are really huge so just switching them of for now
ggplot(data_long, aes(x = Timeperiod, y = Insect_count, fill = Insect_key_name)) + geom_bar(stat = "identity")+ theme_bw()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +ggtitle('Raw abundances') +theme(legend.position = 'none')

ggplot(data_long, aes(x = Timeperiod, y = Insect_count, fill = Insect_key_name)) + geom_bar(stat = "identity")+ theme_bw()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +ggtitle('Raw abundances')+ facet_wrap(vars(Property_name)) +theme(legend.position = 'none')

#relative abundances
data_long_ra<-data_long %>% ungroup() %>%  select(-Tree_number) %>% 
  group_by(Property_name,Timeperiod,Insect_key_name) %>%
  summarise(Insect_count =  sum(Insect_count))%>%  mutate(per =  100 *Insect_count/sum(Insect_count)) %>% 
  ungroup()

ggplot(data_long_ra, aes(x = Timeperiod, y = per, fill = Insect_key_name)) + geom_bar(stat = "identity")+ theme_bw()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +ggtitle('Relative abundances')+ facet_wrap(vars(Property_name)) +theme(legend.position = 'none')
```


## Alpha diversity
For diversity analyses, for now Im just doing by property, can chat with Brad about whether he wants to split it further

```{r alpha, echo=TRUE}
#just going to simplify down to Property_name to start with
data_long_wide_sumproperty<-data_long_wide %>% ungroup() %>% select(-Tree_number,-Timeperiod) %>% group_by(Property_name) %>% summarise(across(everything(), sum,na.rm = TRUE))

#diversity plots
data_long_wide_sumproperty<-data_long_wide_sumproperty %>% remove_rownames %>% column_to_rownames(var="Property_name")

#alpha diversity
#count number of species per property
sppr <- specnumber(data_long_wide_sumproperty)
sppr

#anova against environmental variables, does mean species richness vary among sites?
#not all farms have this plants data:
relavent_farms<-row.names(data_long_wide_sumproperty)
meta_data_farm_sub<-meta_data_farm %>% filter(Property_name %in% relavent_farms)
#try farm.type
sppr_aov <-aov(sppr ~ Farm.type, data = meta_data_farm_sub)
summary(sppr_aov)
#try planting.type
sppr_aov <-aov(sppr ~ Planting.type, data = meta_data_farm_sub)
summary(sppr_aov)
#plant age
sppr_aov <-aov(sppr ~ Plant.age, data = meta_data_farm_sub)
summary(sppr_aov)
#planting area size
sppr_aov <-aov(sppr ~ Planting.area.size, data = meta_data_farm_sub)
summary(sppr_aov)
#no environmental variables significant

#future thing
#should do this across the three different plant species (does diversity change between plant species)

#can plot that out
sppr_df <- sppr %>% 
  enframe() %>% 
  full_join(meta_data_farm_sub, by = c("name" = "Property_name"))

plot_sppr <- ggplot(sppr_df, aes(x = Farm.type, y = value, fill = Farm.type)) +
  geom_boxplot() +
  theme_bw()+
  labs(x = "Farm type",
       y = "Number of species per site",
       title = "Species richness")
plot_sppr
```


## Alpha diversity - shannon
Abundance and eveness

```{r alpha_shannon, echo=TRUE}
#shannon diversity index (takes into account richness and evenness)
#diversity function in vegan
shannondiv <- diversity(data_long_wide_sumproperty)
head(shannondiv)

#anova against shannon diversity, does mean species richness vary among sites?
#try farm.type
sppr_aov <-aov(shannondiv ~ Farm.type, data = meta_data_farm_sub)
summary(sppr_aov)
#try planting.type
sppr_aov <-aov(shannondiv ~ Planting.type, data = meta_data_farm_sub)
summary(sppr_aov)
#plant age
sppr_aov <-aov(shannondiv ~ Plant.age, data = meta_data_farm_sub)
summary(sppr_aov)
#planting area size
sppr_aov <-aov(shannondiv ~ Planting.area.size, data = meta_data_farm_sub)
summary(sppr_aov)

#can plot that out
shannondiv_df <- shannondiv %>% 
  enframe() %>% 
  full_join(meta_data_farm_sub, by = c("name" = "Property_name"))

plot_sppr <- ggplot(shannondiv_df, aes(x = Farm.type, y = value, fill = Farm.type)) +
  geom_boxplot() +
  theme_bw()+
  labs(x = "Farm type",
       y = "Shannon diversity richness",
       title = "Species richness")
plot_sppr
```


## Beta diversity
Using permanova to assess community composition differences

```{r beta, echo=TRUE}
perm <- adonis2(data_long_wide_sumproperty ~ Farm.type, data = meta_data_farm_sub)
perm
perm <- adonis2(data_long_wide_sumproperty ~ Plant.age, data = meta_data_farm_sub)
perm
```

## PCA analysis

```{r pca, echo=TRUE}
#plot them out on PCA 
plantPCA <- rda(data_long_wide_sumproperty)
plantPCA

PCAscores <- scores(plantPCA, display = "sites") %>% 
  as.data.frame() %>% 
  rownames_to_column("site") %>% 
  full_join(meta_data_farm_sub, by = c("site" = "Property_name"))

PCAvect <- scores(plantPCA, display = "species") %>% 
  as.data.frame()


plot_PCA <- ggplot() +
  geom_point(data = PCAscores, aes(x = PC1, y = PC2, color = Farm.type)) +
  labs(title = "Principal Components Analysis") +
  theme_bw()
plot_PCA
#could plot insect species out on them but there is really too many species for this
```


## MDS analysis
I think for the diversity plots it would be good to include the diversity across the three species side by side but will get feedback before going further

```{r mds, echo=TRUE}
plant_NMDS <- metaMDS(data_long_wide_sumproperty)
plant_NMDS
#stress <0.2
stressplot(plant_NMDS)
plot(plant_NMDS)

plot_df <- scores(plant_NMDS, display = "sites") %>% 
  as.data.frame() %>% 
  rownames_to_column("site") %>% 
  full_join(meta_data_farm_sub, by = c("site" = "Property_name"))
plot_nmds <- ggplot(plot_df, aes(x = NMDS1, y = NMDS2, color = Farm.type, shape = Farm.type)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(linetype = 2, size = 1) +
  theme_bw()+
  labs(title = "NMDS")
plot_nmds
```

## MDS with arrows
What species contribute to the structure

```{r mds_arrow, echo=TRUE}
#seeing what species contribute to the plot
# envfit() takes the output of metaMDS() and the species matrix you created
fit <- envfit(plant_NMDS, data_long_wide_sumproperty, perm = 999) 

# extract p-values for each species
fit_pvals <- fit$vectors$pvals %>% 
  as.data.frame() %>% 
  rownames_to_column("species") %>% 
  dplyr::rename("pvals" = ".")

# extract coordinates for species, only keep species with p-val = 0.001
fit_spp <- fit %>% 
  scores(., display = "vectors") %>% 
  as.data.frame() %>% 
  rownames_to_column("species") %>% 
  full_join(., fit_pvals, by = "species") %>% 
  filter(pvals == 0.001)

# new plot
nmds_plot_new <- ggplot(plot_df, aes(x = NMDS1, y = NMDS2)) +
  coord_fixed() +
  geom_point(aes(color = Farm.type, shape = Farm.type), size = 3, alpha = 0.8) +
  stat_ellipse(aes(color = Farm.type)) +
  #scale_color_manual(values = pal) +
  geom_segment(data = fit_spp, aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")),
               col = "black") +
  geom_text(data = fit_spp, aes(label = species)) +
  theme_bw()
nmds_plot_new
```

## MDS and environmental variables
Arrows only work for continuous variables so I cant add in factors

```{r mds_arrow2, echo=TRUE}
ef <- envfit(plant_NMDS ~ Planting.area.size + Plant.age, data = meta_data_farm_sub, perm = 999)
ef

class(meta_data_farm_sub$Planting.type)

# extract p-values for each variable
ef_pvals <- ef$vectors$pvals %>% 
  as.data.frame() %>% 
  rownames_to_column("env_var") %>% 
  dplyr::rename("pvals" = ".")
# extract coordinates for variable, only keep those with p-val = 0.001
ef_spp <- ef %>% 
  scores(., display = "vectors") %>% 
  as.data.frame() %>% 
  rownames_to_column("env_var") %>% 
  full_join(., ef_pvals, by = "env_var") %>% 
  filter(pvals == 0.001)

#none significant so will not show up on plot
# new plot
nmds_plot_new <- ggplot(plot_df, aes(x = NMDS1, y = NMDS2)) +
  coord_fixed() +
  geom_point(aes(color = Farm.type, shape = Farm.type), size = 3, alpha = 0.8) +
  stat_ellipse(aes(color = Farm.type)) +
  #scale_color_manual(values = pal) +
  geom_segment(data = ef_spp, aes(x = 0, xend = NMDS1, y = 0, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")),
               col = "black") +
  geom_text(data = ef_spp, aes(label = env_var)) +
  theme_bw()
nmds_plot_new
```

## Canonical corrrespondence analysis
Im not very familiar with these but it seems to you can have factors in them too...?
With the MDS stuff I was limiting it to significant variables (there was not significant env and only one sig insect in this set but I dont know how to do that for CCA yet)

```{r cca, echo=TRUE}
#think maybe I can add in categorical variables here, lets try
plantCCA <- cca(data_long_wide_sumproperty ~ Farm.type +Planting.area.size + Plant.age+ Planting.type , data = meta_data_farm_sub)
plantCCA
plot(plantCCA)

ccavectors <- as.matrix(scores(plantCCA, display = "bp", scaling = "species")*7.627807) %>% 
  as.data.frame()

# site coordinates
site_data <- scores(plantCCA, display = "sites") %>% 
  as.data.frame() %>% 
  rownames_to_column("site") %>% 
  full_join(meta_data_farm_sub, by = c("site" = "Property_name"))

# species coordinates
species_data <- scores(plantCCA, display = "species") %>% 
  as.data.frame()

# plotting
plot_cca <- ggplot(site_data) +
  geom_point(aes(x = CCA1, y = CCA2, color = Farm.type), shape = 19, size = 2, alpha = 0.8) +
  coord_fixed() +
  #scale_color_manual(values = pal) +
  geom_segment(data = ccavectors, aes(x = 0, y = 0, xend = CCA1, yend = CCA2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
  geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +
#  geom_point(data = species_data, aes(x = CCA1, y = CCA2), shape = 17, size = 2, color = "slateblue") +
#  scale_x_continuous(limits = c(-12, 10)) +
 # scale_y_continuous(limits = c(-3, 12)) +
  geom_text(data = ccavectors, aes(x = CCA1, y = CCA2, label = rownames(ccavectors)), nudge_x = 0.3, nudge_y = 0.3) +
  theme_bw() +
  labs(title = "Canonical Correspondence Analysis")
plot_cca
```

